---
title: "EDSValidation_DataPreparation"
author: "Jennie An"
date: "2025-03-18"
output: html_document
---

# Preparing the data for the validation of Epic's Sepsis Model
## Author: C.H.M. Maas

```{r}
rm(list = ls(all.names = TRUE))
if(!is.null(dev.list())) dev.off()
cat('\014')
```

```{r}
# install.packages("ggplot2")
#install.packages("reshape2")
#install.packages("openxlsx")
```

###libraries
```{r}
library(dplyr)
library(ggplot2)
```

Make sure there are three folders: Code, Data, Results to read from and write to
```{r}
file.path <- "C:/Users/jenan/OneDrive - Tufts Medicine/code/EDSValidation_Jennie/"
```

###load data

Predictions (model.score) are generated every 15 minutes
For negative patients predictions are pulled from admission until discharge
For positive patients predictions are pulled from admission until sepsis onset
```{r}
original.data <- read.csv(paste0(file.path, "Data/SepsisData(X).csv"))
head(original.data)
```


MRN = patient ID
CSN = hospitalization ID
LCSN = Linked CSN, transfers are not considered separate hospitalizations
```{r}
colnames(original.data) <- c("MRN", "CSN", "LCSN", "sex", "race", "ethnicity", "age", 
                             "admission.date.and.time", 
                             "discharge.date.and.time", "center",
                             "prediction.date.and.time", "model.score",
                             "model.prediction", "sep3.time.zero", "sep3.outcome")
head(original.data)
```

check for duplicates
if the sum is zero there is no duplication so no need to adjust
```{r}
sum(duplicated(original.data))
```

###Add columns to data

Separate date and time
```{r}
patient.data <- cbind(original.data, 
                      reshape2::colsplit(original.data$admission.date.and.time, " ", 
                                         names=c("admission.date", "admission.time")),
                      reshape2::colsplit(original.data$discharge.date.and.time, " ", 
                                         names=c("discharge.date", "discharge.time")),
                      reshape2::colsplit(original.data$prediction.date.and.time, " ", 
                                         names=c("prediction.date", "prediction.time")))
head(patient.data)
```

### Admission inclusion criteria

16 August is the date at which predictions started generating
```{r}
cat("Number of patients admitted before 16 August:", sum(patient.data |> 
                                                           filter(admission.date<"2024-08-16") |> 
                                                           summarize(unique_ids=n_distinct(LCSN)) |> 
                                                           pull(unique_ids)), "\n",
    "Number of patients admitted after 31 October:", sum(patient.data |> 
                                                             filter(admission.date>"2024-10-31") |> 
                                                             summarize(unique_ids=n_distinct(LCSN)) |> 
                                                             pull(unique_ids)), "\n")

patient.data <- patient.data |> 
  filter(admission.date>="2024-08-16" & admission.date<"2024-11-01")
cat(" Number of rows:", nrow(patient.data), "\n",
    "Number of unique hospitalizations:", length(unique(patient.data$LCSN)), "\n",
    "Number of unique individuals:", length(unique(patient.data$MRN)), "\n")
```

### Data Manipulations

set time points and maximum time point
```{r}
patient.data <- patient.data |> 
  group_by(LCSN) |> 
  mutate(time.point=0:(n()-1),
         max.time.point=max(time.point))
```

make outcome column time-dependent
```{r}
patient.data$time.sep3.outcome <- rep(0, nrow(patient.data))
patient.data[which(patient.data$time.point==patient.data$max.time.point & 
                     patient.data$sep3.outcome==1), "time.sep3.outcome"] <- 1
```

look at a patient without the outcome
```{r}
View(patient.data |> filter(MRN==31225483))
```

look at a patient with the outcome
```{r}
View(patient.data |> filter(MRN==31259363))
```

look at a patient who had two hospitalizations one month apart
```{r}
View(patient.data |> filter(MRN==31228576))
```

look at patient who had transfer
```{r}
View(patient.data |> filter(MRN==31819490))
```

Exclude patients that immediately have the outcome at the first prediction
```{r}
cat("There are", sum(patient.data$max.time.point==0 & patient.data$sep3.outcome==1),
    "(",
    round(sum(patient.data$max.time.point==0 & patient.data$sep3.outcome==1)/sum(patient.data$time.sep3.outcome==1)*100, 1), 
    "% of positives)",
    "patients that experience the outcome immediately at the time of the first prediction.\n")
patient.data <- patient.data |> filter(!(max.time.point==0 & sep3.outcome==1))
cat(" Number of rows:", nrow(patient.data), "\n",
    "Number of unique hospitalizations:", length(unique(patient.data$LCSN)), "\n",
    "Number of unique individuals:", length(unique(patient.data$MRN)), "\n")
```

look at patient who had two hospitalizations and the second was immediately sepsis
```{r}
View(original.data |> filter(MRN==31235059))
```

```{r}
View(patient.data |> filter(MRN==31235059))
```

check the dates
```{r}
cat('Admission dates lie between', 
    min(patient.data$admission.date), 'and', 
    max(patient.data$admission.date), '\n',
    'Discharge dates lie between', 
    min(patient.data$discharge.date), 'and', 
    max(patient.data$discharge.date), '\n',
    'Prediction dates lie between', 
    min(patient.data$prediction.date), 'and', 
    max(patient.data$prediction.date), '\n')
```

Check that full hospitalization was recorded for the patient with discharge at 28 January
```{r}
View(original.data[which(original.data$MRN=="34145872"),])
```

Check the hospitalization of the patient with prediction date at 21 January
```{r}
View(original.data[which(original.data$MRN=="35471981"), ])
```

Make predictions between 0 and 100
```{r}
patient.data$model.score <- patient.data$model.score/100
```

Group Center
```{r}
patient.data$center.grouped <- ifelse(patient.data$center=="LOWELL GENERAL HOSPITAL MAIN CAMPUS" | patient.data$center=="LOWELL GENERAL HOSPITAL SAINTS CAMPUS", "Lowell",
                                       ifelse(patient.data$center=="MELROSE WAKEFIELD HOSPITAL", "Melrose",
                                              ifelse(patient.data$center=="TUFTS MEDICAL CENTER", "Tufts",
                                                     ifelse(patient.data$center=="HHF HSPC High Pointe House" | patient.data$center=="LAWRENCE MEMORIAL HOSPITAL", "Other", NA))))
```

N/A Sex
```{r}
patient.data <- patient.data |> mutate(sex = na_if(sex, "Unknown"),
                                         sex = na_if(sex, "X"),
                                         sex = na_if(sex, "Nonbinary"))
```

Rename Race
```{r}
patient.data$race.groups <- ifelse(patient.data$race=="White", "White",
                                   ifelse(patient.data$race=="Asian", "Asian",
                                          ifelse(patient.data$race=="Black or African American", "Black",
                                                 ifelse(patient.data$race=="Other" | 
                                                          patient.data$race=="American Indian or Alaska Native" | 
                                                          patient.data$race=="Native Hawaiian / Pacific Islander", 
                                                        "Other", NA))))
```

Fix acceptable ages
```{r}
patient.data <- patient.data |> mutate(age = na_if(age, 125),
                                       age = na_if(age, 124))
```

create age groups
```{r}
quantiles.age <- quantile(aggregate(patient.data |> select(age), 
                                    list(patient.data$LCSN), 
                                    max)[, "age"], na.rm=TRUE)
young.age <- quantiles.age[2]
old.age <- quantiles.age[4]
patient.data$age.groups <- rep(paste("Between", young.age, "and",
                                      old.age, "years"), 
                               nrow(patient.data))
patient.data$age.groups <- ifelse(patient.data$age<young.age, 
                                  paste("Younger than", young.age, "years"),
                                   ifelse(patient.data$age>old.age, 
                                          paste("Older than", old.age, "years"),
                                          ifelse(is.na(patient.data$age), NA, 
                                                 paste("Between", young.age, "and", old.age, "years"))))
head(patient.data)
```

### Data Checks

for all patients, check admission time and first prediction
```{r}
first.prediction <- original.data |> 
  group_by(LCSN) |> 
  slice(1) |> 
  filter(admission.date.and.time>"2024-08-16" & admission.date.and.time<"2024-09-30") |>
  select(admission.date.and.time, prediction.date.and.time)
diff.admission.first.prediction <- as.numeric(difftime(first.prediction$prediction.date.and.time,
                                                       first.prediction$admission.date.and.time,
                                                      units="hours"))
hist(diff.admission.first.prediction, xlab="Time in hours", 
     main="Histogram of time between admission and first prediction")
```

investigate maximum time between admission and first prediction
```{r}
max.ID.admission <- as.numeric(first.prediction[which(diff.admission.first.prediction==max(diff.admission.first.prediction)), "LCSN"])
cat("Maximum time between admission and first prediction is", 
    round(max(diff.admission.first.prediction), 1), "hours \n")
View(original.data |> filter(LCSN==max.ID.admission) |>
      select(MRN, CSN, LCSN, admission.date.and.time, prediction.date.and.time))
```

for negatives, check the time between last prediction and discharge
```{r}
df.negatives <- patient.data |> filter(sep3.time.zero=="NULL" & time.point==max.time.point)
diff.discharge.last.prediction <- as.numeric(difftime(df.negatives$discharge.date.and.time, 
                                                      df.negatives$prediction.date.and.time,
                                                      units="hours"))
hist(diff.discharge.last.prediction, xlab="Time in hours", 
     main="Histogram of time between last prediction and discharge")
```

investigate maximum time between last prediction and discharge
```{r}
max.ID.discharge <- as.numeric(df.negatives[which(diff.discharge.last.prediction==max(diff.discharge.last.prediction)), "LCSN"])
cat("Maximum time between last prediction and discharge is", 
    round(max(diff.discharge.last.prediction), 1), "hours \n")
View(original.data |> filter(LCSN==max.ID.discharge) |>
        select(MRN, CSN, LCSN, discharge.date.and.time, prediction.date.and.time))
```

for positives, check the time between last prediction and sepsis outcome
```{r}
df.positives <- patient.data |> filter(time.sep3.outcome==1)
diff.outcome.last.prediction <- as.numeric(difftime(df.positives$sep3.time.zero, 
                                                    df.positives$prediction.date.and.time,
                                                    units="hours"))
hist(diff.outcome.last.prediction, xlab="Time in hours", 
     main="Histogram of time between last prediction and outcome")
```

investigate maximum time between last prediction and sepsis outcome
```{r}
max.ID.outcome <- as.numeric(df.positives[which(diff.outcome.last.prediction==max(diff.outcome.last.prediction)), "LCSN"])
cat("Maximum time between last prediction and sepsis outcome is", 
    round(max(diff.outcome.last.prediction), 1), "hours \n")
View(original.data |> filter(LCSN==max.ID.outcome) |>
        select(MRN, CSN, LCSN, sep3.time.zero, prediction.date.and.time))
```

missing predictions
```{r}
cat(" For", sum(diff.admission.first.prediction>2), "(",
    round(sum(diff.admission.first.prediction>2)/length(unique(patient.data$LCSN))*100, 1), "%)",
    "patients the time between admission and their first prediction is more than 2 hours. \n",
    "For", sum(diff.discharge.last.prediction>1), "(",
    round(sum(diff.discharge.last.prediction>1)/nrow(df.negatives)*100, 1), "%)",
    "negative patients the time between discharge and the last prediction is more than 1 hour. \n",
    "For", sum(diff.outcome.last.prediction>1), "(",  
    round(sum(diff.outcome.last.prediction>1)/nrow(df.positives)*100, 1), "%)",
    "positive patients the time between sepsis and the last prediction is more than 1 hour. \n")
```

exclude patients with missing predictions
```{r}
IDs <- c(c(first.prediction[diff.admission.first.prediction>2,"LCSN"])$LCSN,
         c(df.positives[diff.outcome.last.prediction>1, "LCSN"])$LCSN,
         c(df.negatives[diff.discharge.last.prediction>1, "LCSN"])$LCSN)
patient.data <- patient.data |> filter(!(LCSN %in% IDs))
cat(" Number of rows:", nrow(patient.data), "\n",
    "Number of unique hospitalizations:", length(unique(patient.data$LCSN)), "\n",
    "Number of unique individuals:", length(unique(patient.data$MRN)), "\n")
```

### INVESTIGATE HOSPITALIZATION TIME

```{r}
times <- patient.data |> filter(max.time.point==time.point) |> 
  select(admission.date.and.time,
         discharge.date.and.time,
         time.sep3.outcome) |>
  mutate(hosp.time=as.numeric(difftime(discharge.date.and.time,
                                       admission.date.and.time,
                                       units="hours")))
hist(times$hosp.time, 
     xlab="Time in hours", ylab="Frequency", breaks=500, 
     main="Time of hospitalization")
```

report hospitalization time
```{r}
cat(" Number of rows:", nrow(patient.data), "\n",
    "Number of unique hospitalizations:", length(unique(patient.data$LCSN)), "\n",
    "Number of unique individuals:", length(unique(patient.data$MRN)), "\n",
    "Average time of hospitalization:", round(mean(times$hosp.time), 1), "hours \n",
    "Average time of hospitalization for negative patients:", 
    round(mean(times$hosp.time[times$time.sep3.outcome==0]), 1), "hours \n",
    "Average time of hospitalization for positive patients:", 
    round(mean(times$hosp.time[times$time.sep3.outcome==1])/24, 1), "days \n",
    "Median time of hospitalization:", round(median(times$hosp.time), 1), "hours \n",
    "Median time of hospitalization for negative patients:", 
    round(median(times$hosp.time[times$time.sep3.outcome==0]), 1), "hours \n",
    "Median time of hospitalization for positive patients:", 
    round(median(times$hosp.time[times$time.sep3.outcome==1])/24, 1), "days \n",
    "Maximum time of hospitalization for positive patients:", 
    round(max(times$hosp.time[times$time.sep3.outcome==1])/24, 1), "days \n")
```

```{r}
# add IQR
cat(
  "Number of rows:", nrow(patient.data), "\n",
  "Number of unique hospitalizations:", length(unique(patient.data$LCSN)), "\n",
  "Number of unique individuals:", length(unique(patient.data$MRN)), "\n",
  "Average time of hospitalization:", round(mean(times$hosp.time), 1), "hours \n",
  "Average time of hospitalization for negative patients:", 
  round(mean(times$hosp.time[times$time.sep3.outcome==0]), 1), "hours \n",
  "Average time of hospitalization for positive patients:", 
  round(mean(times$hosp.time[times$time.sep3.outcome==1])/24, 1), "days \n",
  "Median time of hospitalization:", round(median(times$hosp.time), 1), "hours \n",
  "Q1-Q3 of hospitalization time:", 
  paste0(round(quantile(times$hosp.time, 0.25), 1), "-", 
         round(quantile(times$hosp.time, 0.75), 1)), " hours \n",
  "Median time of hospitalization for negative patients:", 
  round(median(times$hosp.time[times$time.sep3.outcome==0]), 1), "hours \n",
  "Q1-Q3 for negative patients:", 
  paste0(round(quantile(times$hosp.time[times$time.sep3.outcome==0], 0.25), 1), "-",
         round(quantile(times$hosp.time[times$time.sep3.outcome==0], 0.75), 1)), " hours \n",
  "Median time of hospitalization for positive patients:", 
  round(median(times$hosp.time[times$time.sep3.outcome==1])/24, 1), "days \n",
  "Q1-Q3 for positive patients:", 
  paste0(round(quantile(times$hosp.time[times$time.sep3.outcome==1], 0.25)/24, 1), "-",
         round(quantile(times$hosp.time[times$time.sep3.outcome==1], 0.75)/24, 1)), " days \n",
  "Maximum time of hospitalization for positive patients:", 
  round(max(times$hosp.time[times$time.sep3.outcome==1])/24, 1), "days \n"
)
```

prevalence of sepsis
```{r}
times.sepsis <- patient.data |> filter(time.sep3.outcome==1) |>
  pull(max.time.point)/4/24
N.positives <- length(times.sepsis)
hist(times.sepsis, 
     xlab="Time in days", ylab="Frequency", breaks=500, 
     main="Times at which sepsis occurred")
cases.sepsis <- data.frame(times=c("0-12", "12-24", 
                                   "second day",
                                   "third day", 
                                   "fourth day",
                                   "fifth day", 
                                   "sixth day",
                                   "more than six days"),
                           N=c(sum(times.sepsis<0.5),
                               sum(times.sepsis>=0.5 & times.sepsis<1),
                               sum(times.sepsis>=1 & times.sepsis<2),
                               sum(times.sepsis>=2 & times.sepsis<3),
                               sum(times.sepsis>=3 & times.sepsis<4),
                               sum(times.sepsis>=4 & times.sepsis<5),
                               sum(times.sepsis>=5 & times.sepsis<6),
                               sum(times.sepsis>=6)),
                           perc=c(round(sum(times.sepsis<0.5)/N.positives*100, 1),
                                  round(sum(times.sepsis>=0.5 & times.sepsis<1)/N.positives*100, 1),
                                  round(sum(times.sepsis>=1 & times.sepsis<2)/N.positives*100, 1),
                                  round(sum(times.sepsis>=2 & times.sepsis<3)/N.positives*100, 1),
                                  round(sum(times.sepsis>=3 & times.sepsis<4)/N.positives*100, 1),
                                  round(sum(times.sepsis>=4 & times.sepsis<5)/N.positives*100, 1),
                                  round(sum(times.sepsis>=5 & times.sepsis<6)/N.positives*100, 1),
                                  round(sum(times.sepsis>=6)/N.positives*100, 1)))
```

```{r}
cases.sepsis
```

save
```{r}
openxlsx::write.xlsx(data.frame(cases.sepsis), 
           paste0(file.path, "Results/Cases sepsis.xlsx"), 
           rowNames=FALSE)

save(young.age, old.age, patient.data, file=paste0(file.path, "Data/patient.data.RData"))
```













