---
title: "EDSValidation_Epic"
author: "Jennie An"
date: "2025-03-19"
output: html_document
---

#Validation of Epic's Sepsis Model using Epic's 8-hour time window
## Author: C.H.M. Maas

clean environment
```{r}
rm(list = ls(all.names = TRUE))
if(!is.null(dev.list())) dev.off()
cat('\014')
```

Make sure there are three folders: Code, Data, Results to read from and write to
```{r}
file.path <- "C:/Users/jenan/OneDrive - Tufts Medicine/code/EDSValidation_Jennie/"
```

Install PredictionTools: https://github.com/CHMMaas/PredictionTools
```{r}
remotes::install_github("CHMMaas/PredictionTools")
```

libraries
```{r}
library(dplyr)
library(ggplot2)
library(PredictionTools)
rmarkdown::render(paste0(file.path, "Code/HelperFunctions.Rmd"))
```

### Load data

```{r}
load(paste0(file.path, "Data/patient.data.RData"))
```

### Sensitivity Analysis: Sepsis-free Survival

```{r}
#patient.data <- patient.data |>
#   group_by(LCSN) |>
#   filter(time.point >= 12*4)
```

8-hour time window for patients with the outcome
Negative patients: full hospitalization
Positive patients: only those predictions 8 hours before sepsis onset
```{r}
patient.data.8h <- patient.data |> 
  group_by(LCSN) |> 
  filter(sep3.outcome==0 | (sep3.outcome==1 & time.point >= max.time.point - 8*4))
```

look at a patient without the outcome
```{r}
View(patient.data.8h |> filter(MRN==31226594))
```

look at a patient with the outcome
```{r}
View(patient.data.8h |> filter(MRN==31252158))
```

Aggregate data
```{r}
aggregated.data <- aggregate(patient.data.8h, list(patient.data.8h$LCSN), max)
```

###Descriptive statistics

center
```{r}
aggregated.data$center.grouped <- as.factor(aggregated.data$center.grouped)
table(aggregated.data$center.grouped)
cat("Number of unknown center:", sum(is.na(aggregated.data$center.grouped)), "\n")
```

sex
```{r}
table(aggregated.data$sex)
round(table(aggregated.data$sex)/nrow(aggregated.data)*100, 0)
cat("Number of unknown sex:", sum(is.na(aggregated.data$sex)), "\n")
```

age
```{r}
cat("Maximum age:", max(aggregated.data$age, na.rm=TRUE), "\n")
hist(aggregated.data$age, xlab="Age", 
     main="Histogram of age (in years)")
quantile(aggregated.data$age, na.rm=TRUE)
cat("Number of unknown age:", sum(is.na(aggregated.data$age.groups)), "\n")
```

race
```{r}
sort(table(aggregated.data$race.groups))
cat("Number of unknown race:", sum(is.na(aggregated.data$race.groups)), "\n")
```

report outcomes
```{r}
cat("Number of events", sum(patient.data.8h$time.sep3.outcome), "(", 
    round(sum(patient.data.8h$time.sep3.outcome)/nrow(aggregated.data)*100, 1), "%) \n")
hist(patient.data$model.score/100, xlab="Risk score", 
     main="Histogram of risk scores of full data set")
hist(aggregated.data$model.score, xlab="Risk score", 
     main="Histogram of risk scores of aggregated data set (maximum prediction)")
round(quantile(aggregated.data$model.score)*100, 0)
```

calculate lead time: time from alarm until sepsis onset
obtain times at which outcomes occur
results in dataframe with one row per hospitalization and time at which sepsis occurred
```{r}
outcomes <- patient.data.8h[patient.data.8h$time.sep3.outcome==1, 
                            c("LCSN", "time.point")]
```

plot median lead time across range of thresholds
```{r}
plot.lead.times <- c()
threshold.range <- 0:80
table.lead.times.selected <- c()
# calculate lead time in hours for different thresholds
for (try.threshold in threshold.range){
  # indicate when alarm went off by new variable 'alarm'
  patient.data.lead.time <- patient.data.8h |> 
    group_by(LCSN) |> 
    mutate(alarm=as.numeric(model.score>try.threshold/100))
  
  # at each threshold, select alarms and the time at which that alarm occured
  alarms <- patient.data.lead.time[patient.data.lead.time$alarm==1, 
                                   c("LCSN", "time.point")]
  
  # at each threshold, select only first alarm and the time at which first alarm occured
  first.alarms <- alarms |> 
    group_by(LCSN) |> 
    slice(1)
  
  # join two tables
  joined <- right_join(first.alarms, outcomes, by="LCSN")
  colnames(joined) <- c("LCSN", "time.first.alarm", "time.outcome")
  
  # only keep those patients that had an alarm before the outcome 
  joined <- joined |> filter(!is.na(time.first.alarm))
  
  # calculate lead times in hours
  lead.times <- (joined$time.outcome-joined$time.first.alarm)/4
  
  # plot median lead times
  plot.lead.times <- c(plot.lead.times, median(lead.times)) 
  
  if (try.threshold==25 | try.threshold==28 | try.threshold==30 | try.threshold==32){
    table.lead.times.selected <- cbind(table.lead.times.selected,
                                       c(try.threshold,
                                         median(lead.times),
                                         sum(nrow(joined)),
                                         sum(nrow(outcomes)-nrow(joined))))
  }
}
```

```{r}
# with IQR
plot.lead.times <- c()
lead.times.q1 <- c()
lead.times.q3 <- c()
threshold.range <- 0:80
table.lead.times.selected <- c()

# calculate lead time in hours for different thresholds
for (try.threshold in threshold.range) {
  # indicate when alarm went off by new variable 'alarm'
  patient.data.lead.time <- patient.data.8h |> 
    group_by(LCSN) |> 
    mutate(alarm = as.numeric(model.score > try.threshold / 100))
  
  # select alarms and the time at which that alarm occurred
  alarms <- patient.data.lead.time[patient.data.lead.time$alarm == 1, 
                                   c("LCSN", "time.point")]
  
  # select only first alarm for each patient
  first.alarms <- alarms |> 
    group_by(LCSN) |> 
    slice(1)
  
  # join with outcomes
  joined <- right_join(first.alarms, outcomes, by = "LCSN")
  colnames(joined) <- c("LCSN", "time.first.alarm", "time.outcome")
  
  # keep only patients with an alarm before the outcome
  joined <- joined |> filter(!is.na(time.first.alarm))
  
  # calculate lead times in hours
  lead.times <- (joined$time.outcome - joined$time.first.alarm) / 4
  
  # store median, Q1, and Q3
  plot.lead.times <- c(plot.lead.times, median(lead.times))
  lead.times.q1 <- c(lead.times.q1, quantile(lead.times, 0.25))
  lead.times.q3 <- c(lead.times.q3, quantile(lead.times, 0.75))
  
  if (try.threshold %in% c(25, 28, 30, 32)) {
    table.lead.times.selected <- cbind(
      table.lead.times.selected,
      c(try.threshold,
        median(lead.times),
        quantile(lead.times, 0.25),
        quantile(lead.times, 0.75),
        nrow(joined),
        nrow(outcomes) - nrow(joined))
    )
  }
}

# Optionally, name the rows for clarity
rownames(table.lead.times.selected) <- c("Threshold", "Median", "Q1", "Q3", "PatientsWithAlarm", "PatientsWithoutAlarm")
```

```{r}
dim(table.lead.times.selected)
```

```{r}
table.lead.times <- data.frame(threshold=threshold.range, lead.time=plot.lead.times)
rownames(table.lead.times.selected) <- c("Threshold", 
                                         "Lead time",
                                         "Q1",
                                         "Q3",
                                         "Number of patients with alarm before sepsis",
                                         "Number of patients without alarm before sepsis")
openxlsx::write.xlsx(data.frame(table.lead.times.selected), 
                     rowNames=TRUE,
                     file=paste0(file.path, "Results/Lead times across selected thresholds.xlsx"))
```

```{r}
# with IQR
# Create full table with median, Q1, and Q3 for all thresholds
table.lead.times <- data.frame(
  threshold = threshold.range,
  lead.time = plot.lead.times,
  Q1 = lead.times.q1,
  Q3 = lead.times.q3
)

# Add Q1 and Q3 to selected thresholds table
# Make sure table.lead.times.selected includes median, Q1, Q3, patients with/without alarm
table.lead.times.selected <- t(table.lead.times.selected)  # transpose to get thresholds as rows
colnames(table.lead.times.selected) <- c("Threshold", "Median", "Q1", "Q3", 
                                        "PatientsWithAlarm", "PatientsWithoutAlarm")

# Write to Excel
openxlsx::write.xlsx(
  data.frame(table.lead.times.selected),
  rowNames = TRUE,
  file = paste0(file.path, "Results/Lead times across selected thresholds with IQR.xlsx")
)
```

```{r}
ggsave(filename = paste0(file.path, "Results/Lead times.png"), 
       plot = ggplot(table.lead.times, 
                     aes(x = threshold.range, y = lead.time)) +
         geom_line() + 
         geom_hline(yintercept = 1, color = "red") + 
         ylim(0, ceiling(max(table.lead.times$lead.time))) +
         labs(x = "Threshold", y = "Median lead time") +
         theme_bw() +
         theme(text=element_text(size=20)), 
       width = 5, height = 5, units = "in", dpi = 300)
```

report median lead time at a threshold of 25
```{r}
cat(" Median lead time at threshold of 25 is", 
    as.numeric(table.lead.times |> filter(threshold==25) |> select(lead.time)), 
    "hours. \n",
    "Median lead time at threshold of 28 is", 
    as.numeric(table.lead.times |> filter(threshold==28) |> select(lead.time)), 
    "hours. \n",
    "Median lead time at threshold of 30 is", 
    as.numeric(table.lead.times |> filter(threshold==30) |> select(lead.time)), 
    "hours. \n",
    "Median lead time at threshold of 32 is", 
    as.numeric(table.lead.times |> filter(threshold==32) |> select(lead.time)), 
    "hours. \n")
```

Decision-curve analysis
```{r}
grDevices::png(file=paste0(file.path, "Results/DCA.png"),
               width=500, height=500, units="px")
dcurves::dca(sep3.outcome ~ model.score, 
             data=aggregated.data) |>
  plot(smooth=TRUE)
grDevices::dev.off()
```

Obtain results for 4 different thresholds
```{r}
final.thresholds <- c(25, 28, 30, 32)
alarm.fatigue <- c()
compare.thresholds <- c()
compare.thresholds.Tufts <- c()
for (final.threshold in final.thresholds){
  ###
  ### Average number of alarms in September
  ###
  # add alarms to patient data
  patient.data.alarm.fatigue <- patient.data.8h |> 
    group_by(LCSN) |> 
    mutate(alarm=as.numeric(model.score>final.threshold/100))
  
  # select Tufts patients in September
  Tufts.sep.patients <- patient.data.alarm.fatigue |>
    filter(center=="TUFTS MEDICAL CENTER") |>
    filter(prediction.date>="2024-09-01" & prediction.date<="2024-09-30")
  
  # report
  alarm.fatigue <- rbind(alarm.fatigue,
                         c(sum(Tufts.sep.patients$alarm),
                           sum(Tufts.sep.patients$alarm & Tufts.sep.patients$sep3.outcome==1),
                           sum(Tufts.sep.patients$alarm & Tufts.sep.patients$sep3.outcome==0),
                           round(sum(Tufts.sep.patients$alarm)/30, 0),
                           round(sum(Tufts.sep.patients$alarm & Tufts.sep.patients$sep3.outcome==1)/30, 0),
                           round(sum(Tufts.sep.patients$alarm & Tufts.sep.patients$sep3.outcome==0)/30, 0)))
  
  ###
  ### Plot measures for different thresholds
  ###
  nboot <- 10 # FINAL: nboot = 1000
  measures.table <- data.frame(names=c("n", "p", "sensitivity", "specificity",
                                       "PPV", "NPV", "Events", "Cindex", "Cindex_q1", "Cindex_q3", "Eavg", "E90"))
  subgroups <- c("All",  "Male", "Female", "Tufts", "Melrose", "Lowell",
                 paste("Younger than", young.age, "years"),
                 paste("Between", young.age, "and", old.age, "years"),
                 paste("Older than", old.age, "years"),
                 "White", "Asian", "Black", "Other")
  for (subgroup in subgroups){
    # select the relevant hospitalizations
    if (subgroup=="All"){
      # Note: we're using the maximum prediction for each hospitalization
      data.subgroup <- aggregated.data
    } else if (subgroup=="Male" | subgroup=="Female"){
      data.subgroup <- aggregated.data |> filter(sex==subgroup)
    } else if (subgroup=="Tufts" | subgroup=="Melrose" | subgroup=="Lowell"){
      data.subgroup <- aggregated.data |> filter(center.grouped==subgroup)
    } else if (subgroup==paste("Younger than", young.age, "years") |
               subgroup==paste("Between", young.age, "and", old.age, "years") |
               subgroup==paste("Older than", old.age, "years")){
      data.subgroup <- aggregated.data |> filter(age.groups==subgroup)
    } else if (subgroup=="White" | subgroup=="Asian" | subgroup=="Black" |
               subgroup=="Other" | subgroup=="Unknown"){
      data.subgroup <- aggregated.data |> filter(race.groups==subgroup)
    }
    
    # calculate measures
    # print boostrapping lengths
    #print(nrow(data.frame(predictions=data.subgroup$model.score, labels=data.subgroup$sep3.outcome)))
    measures.CI <- measures.with.CI(predictions=data.subgroup$model.score,
                                    labels=data.subgroup$sep3.outcome,
                                    nboot=nboot,
                                    min.pred=0)
    # make a figure of measures
    create.figure(measures=measures.CI, 
                  threshold=final.threshold/100,
                  predictions=data.subgroup$model.score,
                  min.pred=0,
                  title=subgroup,
                  file.path=paste0(file.path, "Results/Threshold ", final.threshold, "/"))
    
    # calibration plot
    calibrated.predictions <- glm(sep3.outcome ~ model.score,
                                  data=data.subgroup,
                                  family="binomial")
    lp <- predict(calibrated.predictions)
    grDevices::png(file=paste0(file.path, "Results/Threshold ", final.threshold, 
                               "/Calibration ", subgroup, ".png"),
                   width=500, height=500, units="px")
    val.prob <- PredictionTools::val.prob.mi(lp.mi=lp, 
                                             lim=c(0, 0.2),
                                             y=data.subgroup$sep3.outcome,
                                             g=ifelse(subgroup=="All", 10, 5), 
                                             main=subgroup, dist=TRUE)
    dev.off()
    
    # calibration plot using raw scores
    if (subgroup=="All"){
      grDevices::png(file=paste0(file.path, "Results/Threshold ", final.threshold, 
                                 "/Calibration ", subgroup, " raw scores.png"),
                     width=500, height=500, units="px")
      rms::val.prob(p=data.subgroup$model.score, 
                    y=data.subgroup$sep3.outcome)
      dev.off()
    }
    
    # add to table
    measures.table <- cbind(measures.table,
                            c(print.measures(threshold=final.threshold/100,
                                             predictions=data.subgroup$model.score,
                                             labels=data.subgroup$sep3.outcome),
                              sprintf("%.3f", val.prob$cindex),
                              sprintf("%.3f", val.prob$cindex.lower),
                              sprintf("%.3f", val.prob$cindex.upper),
                              sprintf("%.3f", val.prob$e.avg),
                              sprintf("%.3f", val.prob$e.90)))
    
    # add to comparison of thresholds
    if (subgroup=="All"){
      compare.thresholds <- cbind(compare.thresholds,
                                c(print.measures(threshold=final.threshold/100,
                                                 predictions=data.subgroup$model.score,
                                                 labels=data.subgroup$sep3.outcome),
                                  sprintf("%.3f", val.prob$cindex)))
    } else if (subgroup=="Tufts"){
      compare.thresholds.Tufts <- cbind(compare.thresholds.Tufts,
                                        c(print.measures(threshold=final.threshold/100,
                                                         predictions=data.subgroup$model.score,
                                                         labels=data.subgroup$sep3.outcome),
                                          sprintf("%.3f", val.prob$cindex)))
    }
  }
  colnames(measures.table) <- c("", subgroups)
  openxlsx::write.xlsx(measures.table, file=paste0(file.path, "Results/Threshold ", 
                                                   final.threshold, "/Measures Table.xlsx"))
}
```

save table for alarm fatigue
```{r}
colnames(alarm.fatigue) <- c("average month", 
                             "average month with sepsis", 
                             "average month without sepsis", 
                             "average day", 
                             "average day with sepsis", 
                             "average day without sepsis")
rownames(alarm.fatigue) <- final.thresholds
openxlsx::write.xlsx(data.frame(alarm.fatigue), 
                     rowNames=TRUE,
                     file=paste0(file.path, "Results/Alarm Fatigue.xlsx"))
```

save table comparing thresholds for all patients
```{r}
colnames(compare.thresholds) <- final.thresholds
rownames(compare.thresholds) <- c("n", "p", "sensitivity", "specificity",
                                  "PPV", "NPV", "Events", "Cindex")
openxlsx::write.xlsx(data.frame(compare.thresholds), 
                     rowNames=TRUE,
                     file=paste0(file.path, 
                                 "Results/Compare performance across thresholds.xlsx"))
```

save table comparing thresholds for Tufts patients
```{r}
colnames(compare.thresholds.Tufts) <- final.thresholds
rownames(compare.thresholds.Tufts) <- c("n", "p", "sensitivity", "specificity",
                                  "PPV", "NPV", "Events", "Cindex")
openxlsx::write.xlsx(data.frame(compare.thresholds.Tufts), 
                     rowNames=TRUE,
                     file=paste0(file.path, 
                                 "Results/Compare performance at Tufts across thresholds.xlsx"))
```
