---
title: "Time_Dependent_Analysis"
output: html_document
---

```{r}
# clean environment
rm(list = ls(all.names = TRUE))
if(!is.null(dev.list())) dev.off()
cat('\014')
```

Install PredictionTools: https://github.com/CHMMaas/PredictionTools
```{r}
remotes::install_github("CHMMaas/PredictionTools")
```

###
### Validation of Epic's Sepsis Model 
### using time-dependent measures with refractory period
### Author: C.H.M. Maas
###

```{r}
library(dplyr)
library(caret)     
library(ggplot2)
library(gridExtra) 
library(cowplot)   
file.path <- "C:/Users/jenan/OneDrive - Tufts Medicine/code/EDSValidation_Jennie/"
rmarkdown::render(paste0(file.path, "Code/HelperFunctions.Rmd"))
load(paste0(file.path, "Data/patient.data.RData"))
```

```{r}
###
### 2-hour look back for predictions
### 8-hours look forward for outcome
###
max.days <- 4                  # range in days for time-dependent analysis
final.threshold <- 30 # c(25, 30)  # set final threshold, vary between 25 and 30
look.back <- 2                 # in hours
look.forward <- 8 # c(8, 12)      # in hours, vary between 8 and 12
```

```{r}
# add alarms to data for each threshold
patient.data <- patient.data |>
  mutate(alarm=as.numeric(model.score>final.threshold/100))
```
    
```{r}
cat(" Threshold:    ", final.threshold, "\n",
      "Look-backward:", look.back, "\n",
      "Look-forward: ", look.forward, "\n")
```
```{r}
# include patients that did not have an alarm look.forward hours
# before the current time point

# first iteration
# length of refractory period is defined by look.forward period
# select: reduce number of variables involved to save time
# mutate:
# 1. in first iteration, define old_alarm when model.score exceeds threshold
# 2. define first alarm by checking if the lagged look.forward*4 alarms are all 0
# 3. define muted alarm by muting all the look.forward*4 alarms after the first alarms
# 4. create new alarm by muting the alarms defined in muted_alarms

patient.data <- patient.data |>
  group_by(LCSN) |>
  select(LCSN, prediction.date.and.time, time.point,
         model.score, alarm,
         max.time.point, time.sep3.outcome) |>
  mutate(old_alarm=as.numeric(model.score>final.threshold/100),
         first_alarm=ifelse(max.time.point == 0, 1,
                            ifelse(old_alarm == 1 &
                                     apply(X=sapply(1:(look.forward*4),
                                                    function(i) lag(old_alarm, i, default = 0)),
                                           MARGIN=1, # apply to rows
                                           FUN=function(x) all(x == 0)), 1, 0)),
         muted_alarm=ifelse(max.time.point == 0, 0,
                            ifelse(apply(X=sapply(1:(look.forward*4),
                                                  function(i) lag(first_alarm, i, default = 0)),
                                         MARGIN=1, # apply to rows
                                         FUN=function(x) any(x == 1)), 1, 0)),
         new_alarm=ifelse(muted_alarm==0, old_alarm, 0))

# repeat these steps until old_alarm and new_alarm are the same
while(sum(patient.data$old_alarm-patient.data$new_alarm)!=0){
  patient.data <- patient.data |>
    group_by(LCSN) |>
    mutate(old_alarm=new_alarm,
           first_alarm=ifelse(max.time.point == 0, 1,
                              ifelse(old_alarm == 1 &
                                       apply(X=sapply(1:(look.forward*4),
                                                      function(i) lag(old_alarm, i, default = 0)),
                                             MARGIN=1, # apply to rows
                                             FUN=function(x) all(x == 0)), 1, 0)),
           muted_alarm=ifelse(max.time.point == 0, 0,
                              ifelse(apply(X=sapply(1:(look.forward*4),
                                                    function(i) lag(first_alarm, i, default = 0)),
                                           MARGIN=1, # apply to rows
                                           FUN=function(x) any(x == 1)), 1, 0)),
           new_alarm=ifelse(muted_alarm==0, old_alarm, 0))
}
```






  
  # remove rows with muted alarms
  patient.data <- patient.data |>
    filter(muted_alarm==0)
  
  # update max.time.point, because the last alarms could be muted and removed
  patient.data <- patient.data |> 
    group_by(LCSN) |> 
    mutate(max.time.point=max(time.point))
  
  # save data with muted alarms
  save(patient.data, 
       file=paste0(file.path, "Data/patient.data.look.backward.", look.back, 
                   ".muted.", look.forward, 
                   ".hours.threshold.", final.threshold, ".RData"))  
} else{
  # load muted data patients file
  load(paste0(file.path, "Data/patient.data.look.backward.", look.back, 
              ".muted.", look.forward, 
              ".hours.threshold.", final.threshold, ".RData"))
}
} else{
# dont mute alarms if no refractory period is activated
patient.data$new_alarm <- patient.data$alarm
patient.data$muted_alarm <- 0
}

# at each time-point, calculate metrics
Prevalence <- c()
at.risk <- c()
TPs <- c()
FPs <- c()
TNs <- c()
FNs <- c()
AUC <- c()
Sensitivity <- c()
Specificity <- c()
PPV <- c()
NPV <- c()
time.points <- seq(look.back*4, max.days*24, by=0.25)  # in hours
for (current.time.point in time.points){
# cat("Current time point: ", current.time.point, " hours \n")

# patients need to survive at least a full look-back period to be included in the analysis
if (current.time.point < look.back*4){
  current.time.point <- look.back*4
}

current.hour <- current.time.point*4

# include patients that were not muted at the current time point
not.muted.patients <- patient.data |>
  group_by(LCSN) |>
  filter(current.hour %in% time.point)

# DISCUSS: take into account prediction at the time point?
# extract alarms in look-back period
pred.df <- not.muted.patients |>
  group_by(LCSN) |>
  filter(max.time.point >= current.hour &   # patient is still in hospitalization
           time.point >= max(0, current.hour - look.back*4) & # look back period
           time.point <= current.hour) |>   # until and including current time point
  summarize(aggr.pred = max(new_alarm))

# DISCUSS: take into account outcome at the time point?
# extract outcome in look-forward period
outcomes <- not.muted.patients |>
  group_by(LCSN) |>
  filter(max.time.point >= current.hour &   # patient is still in hospitalization
           time.point >= current.hour &     # from and including current time point
           time.point <= min(max.time.point,        # look forward period, up until end of hospitalization
                             current.hour + look.forward*4)) |>
  summarize(aggr.outcome = max(time.sep3.outcome))

# join predictions and outcomes
joined <- full_join(pred.df, outcomes, by="LCSN")

# extract predictions and labels
labels <- joined$aggr.outcome
predictions <- joined$aggr.pred

# save AUC
if (sum(labels)==0){
  AUC <- c(AUC, 0.5)
} else{
  AUC <- c(AUC, as.numeric(pROC::auc(labels, predictions, quiet=TRUE)))
}

# save other measures
TP <- sum(predictions==1 & labels==1)
FP <- sum(predictions==1 & labels==0)
TN <- sum(predictions==0 & labels==0)
FN <- sum(predictions==0 & labels==1)
Sensitivity <- c(Sensitivity, TP/(TP+FN))
Specificity <- c(Specificity, TN/(TN+FP))
PPV <- c(PPV, TP/(TP+FP))
NPV <- c(NPV, TN/(TN+FN))
at.risk <- c(at.risk, nrow(joined))
TPs <- c(TPs, TP)
FPs <- c(FPs, FP)
TNs <- c(TNs, TN)
FNs <- c(FNs, FN)
Prevalence <- c(Prevalence, (TP+FP)/nrow(joined)*100)
}

# save results in data.frame
results.df <- data.frame(Time = time.points/24, # in days
                       Prevalence = Prevalence,
                       Risk = at.risk,
                       TP = TPs,
                       FP = FPs,
                       TN = TNs,
                       FN = FNs,
                       AUC = AUC,
                       Sensitivity = Sensitivity,
                       Specificity = Specificity,
                       PPV = PPV,
                       NPV = NPV)

# risk table
table_df <- results.df |>
filter(Time %in% c(look.back*4/24, seq(0.5, max.days, by=0.5))) |>
select(Time, Prevalence, Risk, TP, FP, TN, FN)

risk_table <- ggplot(table_df,
                   aes(x=Time, y=0)) +
geom_text(aes(y=5, label=sprintf("%.1f", Prevalence)), size=3) +
geom_text(aes(y=4, label=Risk), size=3) +
geom_text(aes(y=3, label=TP), size=3) +
geom_text(aes(y=2, label=FP), size=3) +
geom_text(aes(y=1, label=TN), size=3) +
geom_text(aes(y=0, label=FN), size=3) +
scale_x_continuous(breaks = c(round(look.back*4/24, 1), 
                              seq(from=0.5, to=max.days, by=0.5))) +
scale_y_continuous(breaks = 0:5,
                   labels = c("FN", "TN", "FP", "TP", 
                              "# at risk", "Prevalence")) +
labs(title=NULL,
     y=NULL,
     x="Time (in days)", y=NULL) +
theme(plot.background = element_rect(fill="white", linewidth=0),
      panel.background = element_blank(),
      panel.grid = element_blank(),
      axis.line.x = element_line(),
      axis.title.y = element_blank(),
      axis.text.y = element_text(size=10, angle=0, vjust=0.6),
      axis.ticks.y = element_blank())

# plots
plot_list <- list()
for (measure in c("Prevalence", "AUC",
                "Sensitivity", "Specificity",
                "PPV", "NPV")){
# extract measure
results.measure <- results.df[, c("Time", measure)]
colnames(results.measure) <- c("Time", "Measure")

# make plot for measure
plot_measure <- ggplot(results.measure,
                       aes(x = Time, y = Measure)) +
  geom_line(color = "black", linewidth = 0.5) +
  scale_x_continuous(breaks = seq(0, max.days, 1)) + # set x-axis ticks
  labs(title = paste(measure, "calculated at each hour, average:", 
                     sprintf("%.2f", mean(results.measure$Measure))),
       x = NULL,
       y = measure) +
  theme_minimal() +
  theme(plot.background = element_rect(fill="white", linewidth=0, color="white"),
        axis.text = element_text(size = 10))

# adjust y-axis
if (measure=="AUC"){
  plot_measure <- plot_measure +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +  # minimum AUC
    scale_y_continuous(limits=c(0.45, 1),
                       breaks=seq(0.5, 1, 0.1))
} else if (measure=="Prevalence"){
  plot_measure <- plot_measure +
    scale_y_continuous(limits=c(0, ceiling(max(table_df$Prevalence))))
} else{
  plot_measure <- plot_measure + ylim(0, 1)
}

# save plot
plot_list[[measure]] <- plot_measure
}

# save plots
ggsave(paste0(file.path, "Results/Threshold ", final.threshold,
            "/Time dependent measures",
            ifelse(refractory, " muting alarms", ""), " look back ",
            look.back, " and look forward ", look.forward, ".png"),
     plot = ggpubr::ggarrange(plot_list[["Prevalence"]],
                              plot_list[["AUC"]],
                              plot_list[["Sensitivity"]],
                              plot_list[["Specificity"]]+
                                scale_y_continuous(limits=c(0.75, 1)),
                              plot_list[["PPV"]]+
                                scale_y_continuous(limits=c(0, 0.25)),
                              plot_list[["NPV"]]+
                                scale_y_continuous(limits=c(0.9, 1)),
                              risk_table, nrow=7, ncol = 1, align="hv"),
     width = 10, height = 12, dpi = 300)
# show patient at time-point 384 and onward to see it's going well
# View(patient.data |> filter(LCSN==3100532557))